---
title: "03-Survival-analysis-regression-models"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = param, eval= param, include=T)
```

### Quick tasks

Consider the `Melanoma` data set.

- Make Kaplan-Meier curves using the same `Surv` object defined above, i.e. with both censoring categories combined into a single one. Do this both for the entire data together, as well as for the observations according to gender.

```{r}
survm <- Surv(Melanoma$time, Melanoma$status == 1)
```

```{r}
plot(survfit(survm ~ 1))
```

```{r}
plot(survfit(survm ~ Melanoma$sex))
```

- Compute the log-rank test to check whether there is evidence that the two survival probability curves are different.

```{r}
survdiff(survm ~ Melanoma$sex)
```

Let us consider again the `leuk` example with overall survival data of leukemia cases, of the package `MASS`. 

```{r}
library(MASS)
library(survival)
str(leuk)
```

We define again the response as a function of `time` only, since all patients experience an event (death). Subsequently, we fit a survival regression of the survival response as a function of `ag` and of white blood cell count `wbc`. We will use Cox regression for this.

- First, let us examine the nature of the relationship between `time` and the covariates `ag` and `wbc`. For this, we `log(time)` against each covariate. We are plotting `time` on the log-scale, as this is equivalent to what happens in the regression model.

```{r, fig.width = 10, fig.height = 5}
par(mfrow = c(1, 2))
plot(log(time) ~ ag + wbc, data = leuk)
```

- The covariate `wbc` has many small values and only a few large ones. This is characteristic of a variable on an exponential scale. So let's take its logarithm and make the plot again.

```{r}
plot(log(time) ~ log(wbc), data = leuk)
```

    This relationship seems to be closer to linear. So, we will include `wbc` in the model with a log-transformation.

- Now let's perform a Cox regression using the function `coxph()`.

    The `coxph()` function, like the regression functions `lm` and `glm`, expect a formula as input. The formula uses a `Surv` object as response on the left hand side, and the covariates on the right hand side (in this case `ag` and `log(wbc)`).

```{r}
leuk.cox <- coxph(Surv(time) ~ ag + log(wbc), data = leuk)
summary(leuk.cox)
```

Consider again the Melanoma data, and now define `sex` as 1 for females and 0 for males.

- Fit the Cox regression model again and compare results with those when `sex` was used with the original coding. 

```{r}
sex1 <- 1 - Melanoma$sex
mel.cox1 <- coxph(survm ~ sex1)
summary(mel.cox1)
# With the original coding we had:
mel.cox <- coxph(Surv(time, status == 1) ~ sex, data = Melanoma)
mel.cox
```
